A Continuation-Based Solution\break of the Linearity Challenge

### 2025-12-29 paper/Congruence.lagda.tex
global revision

There are two non-trivial aspects worth commenting. The first one concerns the
proof $\AgdaInductiveConstructor{>}~q$ used in the
\AgdaInductiveConstructor{wait} process before structural pre-congruence is
applied. To understand the meaning of this proof, we must recall three key
elements:
\begin{enumerate}
  \item
    $(\AgdaInductiveConstructor{wait}~(\AgdaInductiveConstructor{ch~⟨~>}~q~\AgdaInductiveConstructor{⟩}~P)$
    is a direct sub-process of the \AgdaInductiveConstructor{cut}, and therefore
    it is meant to be well typed in the context
    $A~\AgdaInductiveConstructor{∷}~\Context_1$.
  \item Being a \AgdaInductiveConstructor{wait} process, such context must
    contain a $\Bot$ type as per the typing rule \WaitRule. That is
    $\Splitting{A~\AgdaInductiveConstructor{∷}~\Context_1}{[\Bot]}{A~\AgdaInductiveConstructor{∷}~\ContextD}$
    for some $\ContextD$.
  \item The \SWait rule is applicable only provided that the channel restricted
    by the cut (say $x$, of type $A$) is different from the channel  consumed by
    the \AgdaInductiveConstructor{wait} process (say $y$, of type $\Bot$).
    %
    We enforce the side condition $x \ne y$ of \SWait (which we left implicit in
    \Cref{tab:semantics}) imposing that the type $A$ in front of
    $A~\AgdaInductiveConstructor{∷}~\Context_1$ goes to the right partition of
    the splitting $[\Bot] + A~\AgdaInductiveConstructor{∷}~\Delta$ through the
    use of \AgdaInductiveConstructor{>}.
\end{enumerate}

@@

Here the process
$(\AgdaInductiveConstructor{select}~(\AgdaInductiveConstructor{ch ⟨
>}~q~\AgdaInductiveConstructor{⟩})~\AgdaInductiveConstructor{inj₁}~P)$, that is
$\Select\y\LeftTag\z.P$, is found under a cut for $x : A$ and is using some
channel $y : B \Plus C$ to send the tag $\InTag_1$.
%
Unlike \AgdaInductiveConstructor{wait}, which consumes the channel it operates
on, the continuation process $P$ uses a fresh continuation channel $z : B$.
Therefore, $P$ is required to be well typed in the context
$B~\AgdaInductiveConstructor{∷}~A~\AgdaInductiveConstructor{∷}~\ContextD$, where
the type $B$ of the continuation $z$ comes \emph{before} the type $A$ of the
restricted channel $x$.
%
After structural pre-congruence is applied, however, the type of the
continuation channel $z$ ends up behind that of the restricted channel $x$,
because now $z$ and $x$ are introduced in the opposite order. Therefore, we need
to rename the channels in $P$ so that it is well typed in the context
$A~\AgdaInductiveConstructor{∷}~B~\AgdaInductiveConstructor{∷}~\ContextD$. Such
renaming is achieved applying the function \AgdaFunction{↭proc} to the
\AgdaInductiveConstructor{swap} permutation and to the process $P$.

@@

Recall from \Cref{tab:semantics} that we want to apply this rule on a process of
the form $\Cut[A]\x{\Fork\y\u\v{P}{Q}}{R}$ when $x \in \fn{P}$. We capture this
condition by means of the splitting $\AgdaInductiveConstructor{<}~r$, implying
that the type $A$ of $x$ ends up in the typing context for $P$ and not in the
one for $Q$.
%
The symmetric rule \SForkR is modeled by another constructor
\AgdaInductiveConstructor{s-fork-r}, which is similar to
\AgdaInductiveConstructor{s-fork-l} except that $\AgdaInductiveConstructor{<}~r$
is replaced by $\AgdaInductiveConstructor{>}~r$.

@@

We concede that the implementation of the pre-congruence rules (including those
not discussed here) can be difficult to decipher. In part, this is due to the
fact that splitting proofs are manifest and no longer hidden by separating
conjunctions (as in \Cref{sec:process-agda}) because we need them to enforce the
side conditions of the rules in \Cref{tab:semantics}. We should also bear in
mind that, using an intrinsically-typed representation of processes,
pre-congruence and reduction rules (which we describe next) incorporate the
proof that they preserve typing.

>>>>

There are two non-trivial aspects worth commenting. The first one concerns the
proof $\AgdaInductiveConstructor{>}~q$ used in the
\AgdaInductiveConstructor{wait} process. To understand the meaning of this
proof, we must recall three key elements:
\begin{enumerate}
  \item
    $(\AgdaInductiveConstructor{wait}~(\AgdaInductiveConstructor{ch~⟨~>}~q~\AgdaInductiveConstructor{⟩}~P)$
    is a direct sub-process of the \AgdaInductiveConstructor{cut}, and therefore
    it is meant to be well typed in the context
    $A~\AgdaInductiveConstructor{∷}~\Context_1$.
  \item Being a \AgdaInductiveConstructor{wait} process, such context must
    contain a $\Bot$ type as per the typing rule \WaitRule. That is
    $\Splitting{A~\AgdaInductiveConstructor{∷}~\Context_1}{[\Bot]}{A~\AgdaInductiveConstructor{∷}~\ContextD}$
    for some $\ContextD$.
  \item The \SWait rule is applicable only provided that the channel restricted
    by the cut (say $x$, of type $A$) is different from the channel  consumed by
    the \AgdaInductiveConstructor{wait} process (say $y$, of type $\Bot$).
    %
    We enforce the side condition $x \ne y$ of \SWait (which we left implicit in
    \Cref{tab:semantics}) imposing that the type $A$ in front of
    $A~\AgdaInductiveConstructor{∷}~\Context_1$ goes to the right partition of
    the splitting $[\Bot] + A~\AgdaInductiveConstructor{∷}~\Delta$ through the
    use of \AgdaInductiveConstructor{>}.
\end{enumerate}

@@

Here the process
$(\AgdaInductiveConstructor{select}~(\AgdaInductiveConstructor{ch ⟨
>}~q~\AgdaInductiveConstructor{⟩})~\AgdaInductiveConstructor{inj₁}~P)$, that is
$\Select\y\LeftTag\z.P$, is found under a cut for $x : A$ and is using some
channel $y : B \Plus C$ to select $\InTag_1$.
%
The continuation process $P$ uses a fresh continuation channel $z : B$.
Therefore, $P$ is required to be well typed in the context
$B~\AgdaInductiveConstructor{∷}~A~\AgdaInductiveConstructor{∷}~\ContextD$, where
the type $B$ of $z$ comes \emph{before} the type $A$ of $x$ since $z$ is
introduced later than $x$.
%
After structural pre-congruence is applied, however, the type of the
continuation channel $z$ ends up behind that of the restricted channel $x$,
because now $z$ and $x$ are introduced in the opposite order. Therefore, we need
to rename the channels in $P$ so that it is well typed in the context
$A~\AgdaInductiveConstructor{∷}~B~\AgdaInductiveConstructor{∷}~\ContextD$. Such
renaming is achieved applying the function \AgdaFunction{↭proc} to the
\AgdaInductiveConstructor{swap} permutation and to the process $P$.

@@

Recall from \Cref{tab:semantics} that we allow using this rule on a process of
the form $\Cut[A]\x{\Fork\y\u\v{P}{Q}}{R}$ when $x \in \fn{P}$. We capture the
condition $x\in\fn{P}$ by means of the splitting
$\AgdaInductiveConstructor{<}~r$, implying that the type $A$ of $x$ ends up in
the typing context for $P$ and not in the one for $Q$.
%
The symmetric rule \SForkR is modeled by another constructor
\AgdaInductiveConstructor{s-fork-r}, which is similar to
\AgdaInductiveConstructor{s-fork-l} except that $\AgdaInductiveConstructor{<}~r$
is replaced by $\AgdaInductiveConstructor{>}~r$.

@@

We concede that the implementation of the pre-congruence rules (including those
not discussed here) can be difficult to decipher. In part, this is due to the
fact that splitting proofs are manifest and no longer hidden by separating
conjunctions (as in \Cref{sec:process-agda}) because we need them to enforce the
side conditions of the rules in \Cref{tab:semantics}. We should also bear in
mind that, using an intrinsically-typed representation of processes, we have
already taken care of the proof that structural pre-congruence preserves typing.

### 2025-12-29 paper/Reduction.lagda.tex
global revision

The splitting $p$ indicates that $y$ (of type $\dual{A}$) occurs on the left
sub-process of the cut (that is the link $\Link\x\y$) and the splitting
$\AgdaInductiveConstructor{<}~\AgdaInductiveConstructor{>}~\AgdaInductiveConstructor{•}$
to which \AgdaInductiveConstructor{link} is applied is structured consistently
with the syntax of the link $\Link\x\y$ being reduced, which is oriented so that
the restricted channel $x$ is on the left. 
%
The process $P$ has type
$\AgdaDatatype{Proc}~(\AgdaFunction{dual}~A~\AgdaInductiveConstructor{∷}~\ContextD)$
and turns into $P\subst\y\x$ after the reduction. The type
$\AgdaFunction{dual}~A$ of $x$ in $P$ is the first in the typing context,
indicating that it is the newest channel that $P$ is aware of. However, after
the reduction, $x$ is replaced by $y$ which is found \emph{somewhere} within
$\Context$. The exact location of $y$ in $\ContextC$ is encoded in the splitting
$p$, thus we ``rename'' $x$ into $y$ within $P$ using the permutation
$\AgdaFunction{↭concat}~p$.

>>>>

The splitting $p$ indicates that $y$ (of type $\dual{A}$) occurs in the left
sub-process of the cut (that is the link $\Link\x\y$) and the splitting
$\AgdaInductiveConstructor{<}~\AgdaInductiveConstructor{>}~\AgdaInductiveConstructor{•}$
to which \AgdaInductiveConstructor{link} is applied is structured consistently
with the syntax of the link being reduced, which is oriented so that the
restricted channel $x$ is on the left. 
%
The process $P$ has type
$\AgdaDatatype{Proc}~(\AgdaFunction{dual}~A~\AgdaInductiveConstructor{∷}~\ContextD)$
and turns into $P\subst\y\x$ after the reduction. The type
$\AgdaFunction{dual}~A$ of $x$ in $P$ is the first in the typing context,
indicating that it is the newest channel that $P$ is aware of. However, after
the reduction, $x$ is replaced by $y$ which is found \emph{somewhere} within
$\Context$. The exact location of $y$ in $\ContextC$ is encoded in the splitting
$p$, thus we ``rename'' $x$ into $y$ within $P$ using the permutation
$\AgdaFunction{↭concat}~p$.

### 2025-12-29 paper/related.tex
global revision

There is no strong evidence that the intrinsically typed representation of
processes reduces the size of the formalisation. As observed in
\Cref{sec:process-agda}, this choice helps reducing the overall number of
datatypes to be defined and makes some results trivial (\eg
\Cref{thm:type-safety} formalised by \AgdaFunction{type-safety}), but the
definitions are also more involved because they incorporate invariants and the
proofs of typing preservation. We speculate that the effort for representing
processes, types and typing rules is not substantially impacted overall.

>>>>

There is no strong evidence that the intrinsically typed representation of
processes reduces the size of the formalisation. As observed in
\Cref{sec:process-agda}, this choice helps reducing the overall number of
datatypes to be defined and makes some results trivial (\eg
\Cref{thm:type-safety} formalised by \AgdaFunction{type-safety}), but the
definitions are also more involved because they incorporate invariants and bits
of the proofs of typing preservation. We speculate that the effort for
representing processes, types and typing rules is not substantially impacted
overall.
